soa

### 2024-12-12 inst/manuscript_figures.Rmd
most recent updates

library(tidyverse)
library(readxl)
library(ggrepel)
library(pheatmap)

@@

@@

  x <- NULL
  for (coef in colnames(contrasts)) {
    res <- topTable(fit, coef=coef, number=Inf, sort.by="none")
    res <- as_tibble(res, rownames="gene")
    res <- res %>%
      select(gene,
             log2fc=logFC,
             avgexpr=AveExpr,
             pval=P.Value,
             padj=adj.P.Val) %>%
      mutate(de = case_when(padj > padj_cutoff ~ "no",
                            log2fc < -log2fc_cutoff ~ "dn",
                            log2fc > log2fc_cutoff ~ "up",
                            TRUE ~ "no"),
             contrast = coef)
    x <- bind_rows(x, res)
  }
  return(x)
}

@@

```{r de heatmaps, include=FALSE}

# heatmap functions
plot_pheatmap <- function(m, 
                          annot_col,
                          # annot_row, 
                          annot_colors,
                          filename, width, height,
                          xmin=-6, xmax=6) {
  pheatmap(m,
           scale = "none",
           color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
           breaks = seq(xmin, xmax, length.out=51),
           annotation_col = annot_col,
           # annotation_row = annot_row,
           # annotation_colors = annot_colors,
           border_color = NA,
           cluster_rows = TRUE,
           cluster_cols = TRUE,
           clustering_distance_rows = "euclidean",
           clustering_distance_cols = "euclidean",
           clustering_method = "ward.D2",
#           cellwidth = 6,
#           cellheight = 6,
           fontsize = 6,
           fontsize_row = 6,
           fontsize_col = 6,
           filename = filename, 
           width = width,
           height = height)
}

# setup heatmap annotations
# column annotations: show clusters, study
# no row annotations
gene_annot <- fds_qq_qn$meta
# %>%
#   left_join(hpa_pdac, by=c("gene"="gene_symbol")) %>%
#   left_join(cptac_de_merged, by=c("gene"="gene_symbol")) %>%
#   mutate(de_cptac_rna = replace_na(de_cptac_rna, "na"),
#          de_cptac_prot = replace_na(de_cptac_prot, "na"))
# annot_row <- gene_annot %>%
#   select(gene) %>%
#   column_to_rownames("gene") %>%
#   as.data.frame()
annot_col <- column_to_rownames(fds_qq_qn$samples, "aoi") %>%
  select(study, cluster_name) %>%
  as.data.frame()

#
# IPMN clusters
#
# de gene params
padj_cutoff <- 0.01
log2fc_cutoff <- 3
ntop <- 10
contrasts <- c("clusterPanc", "clusterLn", "clusterKidney", "clusterColon", "clusterBrain")

g <- filter(de_qq_qn, contrast %in% contrasts,
            abs(log2fc) > log2fc_cutoff, 
            padj < padj_cutoff) %>%
  group_by(contrast, de) %>%
  slice_max(abs(log2fc), n=ntop) %>%
  ungroup() %>%
  distinct(gene) %>% 
  pull(gene)
ngenes <- length(g)
cluster_marker_genes <- g

# centered gene expression (subtract median)
s <- fds_qq_qn$samples
x <- log2_norm_cpm_qq_qn[g, s$aoi]
u <- apply(x, 1, mean)
x <- sweep(x, 1, u, FUN="-")
xmin <- -6
xmax <- 6

# heatmap
heatmap_dir <- file.path(plot_dir, "heatmap")

f <- file.path(plot_dir,  paste0("heatmap_de_cluster_markers_", ngenes, "_genes"))
h <- 6
w <- 6
p <- plot_pheatmap(x, 
                   annot_col[s$aoi,], 
                   annot_row[g,], 
                   annot_colors=scale_color_continuous(),
                   filename=paste0(f, ".png"), 
                   width=w, height=h,
                   xmin=xmin, xmax=xmax)
save_pheatmap_pdf(p, filename=paste0(f, ".pdf"), width=w, height=h)

@@

for (gene in feature_plot_genes) {
  print(gene)
  x <- bind_cols(s, as_tibble(pca_res_qq_qn$x), gene=unlist(y[gene,]))
  p <- ggplot(x, aes(x=PC1, y=PC2, color=gene, shape=study)) +
    geom_point(alpha = 0.8, size=3) +
    scale_color_viridis_c() +
    theme_minimal() +
    theme(aspect.ratio=1) +
    labs(x="PC1", y="PC2", title=gene)
  f <- file.path(plot_dir, paste0("pca_featureplot_", gene))
  ggsave(paste0(f, ".pdf"), plot=p, width=5, height=5)
  ggsave(paste0(f, ".png"), plot=p, width=5, height=5)
}

@@

plot_de_scatter2d <- function(tbl, xcontrast, ycontrast, nlabel=10, 
                              highlight_genes=NULL) {
  if (is.null(highlight_genes)) {
    highlight_genes <- c()
  }
  xde <- sym(paste0("de_", xcontrast))
  yde <- sym(paste0("de_", ycontrast))
  xlog2fc <- sym(paste0("log2fc_", xcontrast))
  ylog2fc <- sym(paste0("log2fc_", ycontrast))

  x <- tbl %>% 
    mutate(subtype = paste0("x", {{xde}}, "_", "y", {{yde}}),
           log2fc_abs_x = case_when({{xde}} == "no" ~ 0,
                                    TRUE ~ abs({{xlog2fc}})),
           log2fc_abs_y = case_when({{yde}} == "no" ~ 0,
                                    TRUE ~ abs({{ylog2fc}})),
           log2fc_abs_max = pmax(log2fc_abs_x, log2fc_abs_y))

  color_scale = c(xno_yno = "#888888",
                  xup_yup = "#FF0000",
                  xdn_ydn = "#0000aa",
                  xup_ydn = "#eeee00",
                  xdn_yup = "#ff00ff",
                  xup_yno = "#32cd32",
                  xdn_yno = "#670092",
                  xno_yup = "#ff6700",
                  xno_ydn = "#16ccc4")
  size_scale = c(xno_yno = 1, 
                 xup_yup = 2,
                 xdn_ydn = 2,
                 xup_ydn = 2,
                 xdn_yup = 2,
                 xup_yno = 2,
                 xdn_yno = 2,
                 xno_yup = 2,
                 xno_ydn = 2)

  annot_data <- filter(x, subtype != "xno_yno")
  none_data <- filter(x, subtype == "xno_yno")
  label_data <- annot_data %>% 
    group_by(subtype) %>%
    slice_max(log2fc_abs_max, n=nlabel) %>%
    ungroup()
  highlight_data <- filter(x, gene %in% highlight_genes)

  p <- ggplot(x, aes(x={{xlog2fc}}, 
                     y={{ylog2fc}},
                     color=subtype, 
                     size=subtype)) +
    geom_point(data=none_data, alpha=0.4) + 
    geom_point(data=annot_data, alpha=0.8) +
    geom_text_repel(data=label_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) + 
    geom_point(data=highlight_data, pch=21, fill=NA, size=5, colour="#ffff00", stroke=1) + 
    geom_point(data=highlight_data, pch=21, fill=NA, size=4, colour="#222222", stroke=1, alpha=0.5) +
    geom_text_repel(data=highlight_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) +     
    scale_color_manual(values = color_scale) +
    scale_size_manual(values = size_scale, guide = "none") + 
    theme_minimal() +
    xlab(rlang::as_string(xlog2fc)) +
    ylab(rlang::as_string(ylog2fc)) +
    labs(color = "DE Genes")
  return(p)
}

# merge de analysis by gene
de_merged <- de_qq_qn %>% select(gene, contrast, log2fc, padj, de) %>%
  pivot_wider(names_from = contrast, values_from = c(log2fc, padj, de))
nlabel <- 10
width <- 8
height <- 8

#
# IPMN Clusters
#
p <- plot_de_scatter2d(de_merged, 
                       xcontrast="clusterColon", 
                       ycontrast="clusterLn", 
                       nlabel=nlabel) +
  labs(title="IPMN Specimens", 
       subtitle="clusterColon (x-axis) clusterLn (y-axis)")
f <- file.path(plot_dir, "scatter2d_de_x_clusterColon_y_clusterLn")
ggsave(paste0(f, ".pdf"), plot=p, width=8, height=8)
ggsave(paste0(f, ".png"), plot=p, width=8, height=8)

@@

## FIGURE 6: GSEA
``` {r gsea}
gsea_nperms <- 10000

read_msigdb_json_file <- function(filename) {
  x <- jsonlite::fromJSON(filename)
  y <- tibble(gs_name = names(x), gs = x) %>% 
    unnest_wider(gs) %>%
    unnest_longer(geneSymbols) %>%
    select(gs_cat = collection,
           gs_name = gs_name,
           gene_symbol = geneSymbols)
}

# read msigdb data: hallmark
f <- system.file("extdata", "h.all.v2024.1.Hs.json", package = "stgeomx")
msigdb_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
msigdb_genes <- intersect(fds_qq_qn$meta$gene, msigdb_gene_sets$gene_symbol)
# filter genes not in our dataset
msigdb_gene_sets <- msigdb_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_hallmark <- dplyr::filter(msigdb_gene_sets, gs_cat == "H")
gs_hallmark <- split(x = gs_hallmark$gene_symbol, f = gs_hallmark$gs_name)

# read msigdb data: pancreas_endocrine
f <- system.file("extdata", "GOBP_ENDOCRINE_PANCREAS_DEVELOPMENT.v2024.1.Hs.json", package = "stgeomx")
gobp_end_pancreas_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_end_pancreas_genes <- intersect(fds_qq_qn$meta$gene, gobp_end_pancreas_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_end_pancreas_gene_sets <- gobp_end_pancreas_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_end_panc <- dplyr::filter(gobp_end_pancreas_gene_sets, gs_cat == "C5:GO:BP")
gs_end_panc <- split(x = gs_end_panc$gene_symbol, f = gs_end_panc$gs_name)

# read msigdb data: pancreas secretion
f <- system.file("extdata", "GOBP_INSULIN_SECRETION.v2024.1.Hs.json", package = "stgeomx")
gobp_pancreas_secretion_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_pancreas_secretion_genes <- intersect(fds_qq_qn$meta$gene, gobp_pancreas_secretion_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_pancreas_secretion_gene_sets <- gobp_pancreas_secretion_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_panc_sec <- dplyr::filter(gobp_pancreas_secretion_gene_sets, gs_cat == "C5:GO:BP")
gs_panc_sec <- split(x = gs_panc_sec$gene_symbol, f = gs_panc_sec$gs_name)

# read msigdb data: brain development gene sets
f <- system.file("extdata", "brain_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_brain_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_brain_genes <- intersect(fds_qq_qn$meta$gene, gobp_brain_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_brain_gene_sets <- gobp_brain_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_brain <- dplyr::filter(gobp_brain_gene_sets, gs_cat == "C5:GO:BP")
gs_brain <- split(x = gs_brain$gene_symbol, f = gs_brain$gs_name)

# read msigdb data: ln development gene sets
f <- system.file("extdata", "ln_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_ln_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_ln_genes <- intersect(fds_qq_qn$meta$gene, gobp_ln_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_ln_gene_sets <- gobp_ln_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_ln <- dplyr::filter(gobp_ln_gene_sets, gs_cat == "C5:GO:BP")
gs_ln <- split(x = gs_ln$gene_symbol, f = gs_ln$gs_name)

# read msigdb data: kidney development gene sets
f <- system.file("extdata", "kidney_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_kidney_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_kidney_genes <- intersect(fds_qq_qn$meta$gene, gobp_kidney_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_kidney_gene_sets <- gobp_kidney_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_kidney <- dplyr::filter(gobp_kidney_gene_sets, gs_cat == "C5:GO:BP")
gs_kidney <- split(x = gs_kidney$gene_symbol, f = gs_kidney$gs_name)

# read msigdb data: colon development gene sets
f <- system.file("extdata", "colon_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_colon_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_colon_genes <- intersect(fds_qq_qn$meta$gene, gobp_colon_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_colon_gene_sets <- gobp_colon_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_colon <- dplyr::filter(gobp_colon_gene_sets, gs_cat == "C5:GO:BP")
gs_colon <- split(x = gs_colon$gene_symbol, f = gs_colon$gs_name)

# new gsea code
get_fgsea_ranks_de <- function(xde, mycontrast) {
  ranks <- xde %>%
    dplyr::filter(contrast == mycontrast) %>%
    select(gene, log2fc, padj) %>%
    mutate(rank = log2fc)
  ranks = sort(setNames(ranks$rank, ranks$gene), decreasing = TRUE)
  return(ranks)
}

  res <- fgsea_run_batch(de, gs_ex_panc,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_end_panc,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_panc_sec,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_brain,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_ln,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_kidney,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_colon,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

>>>>

library(tidyverse)
library(readxl)
library(ggrepel)
library(pheatmap)
library(dplyr)
library(tidyr)
library(jsonlite)

@@

# load package
install.packages(working_dir, repos = NULL, type = "source")

@@

  x <- NULL
  res <- NULL

  for (coef in colnames(contrasts)) {
    res <- topTable(fit, coef=coef, number=Inf, sort.by="none")
    res <- as_tibble(res, rownames="gene")
    res <- res %>%
      dplyr::select(gene,
             log2fc=logFC,
             avgexpr=AveExpr,
             pval=P.Value,
             padj=adj.P.Val) %>%
      mutate(de = case_when(padj > padj_cutoff ~ "no",
                            log2fc < -log2fc_cutoff ~ "dn",
                            log2fc > log2fc_cutoff ~ "up",
                            TRUE ~ "no"),
             contrast = coef)
    x <- bind_rows(x, res)
  }
  return(x)
}

@@

 ```{r de heatmaps, include=FALSE}
# 
# # heatmap functions
# plot_pheatmap <- function(m, 
#                           annot_col,
#                           # annot_row, 
#                           annot_colors,
#                           filename, width, height,
#                           xmin=-6, xmax=6) {
#   pheatmap(m,
#            scale = "none",
#            color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
#            breaks = seq(xmin, xmax, length.out=51),
#            annotation_col = annot_col,
#            # annotation_row = annot_row,
#            # annotation_colors = annot_colors,
#            border_color = NA,
#            cluster_rows = TRUE,
#            cluster_cols = TRUE,
#            clustering_distance_rows = "euclidean",
#            clustering_distance_cols = "euclidean",
#            clustering_method = "ward.D2",
# #           cellwidth = 6,
# #           cellheight = 6,
#            fontsize = 6,
#            fontsize_row = 6,
#            fontsize_col = 6,
#            filename = filename, 
#            width = width,
#            height = height)
# }
# 
# # setup heatmap annotations
# # column annotations: show clusters, study
# # no row annotations
# gene_annot <- fds_qq_qn$meta
# # %>%
# #   left_join(hpa_pdac, by=c("gene"="gene_symbol")) %>%
# #   left_join(cptac_de_merged, by=c("gene"="gene_symbol")) %>%
# #   mutate(de_cptac_rna = replace_na(de_cptac_rna, "na"),
# #          de_cptac_prot = replace_na(de_cptac_prot, "na"))
# # annot_row <- gene_annot %>%
# #   select(gene) %>%
# #   column_to_rownames("gene") %>%
# #   as.data.frame()
# annot_col <- column_to_rownames(fds_qq_qn$samples, "aoi") %>%
#   select(study, cluster_name) %>%
#   as.data.frame()
# 
# #
# # IPMN clusters
# #
# # de gene params
# padj_cutoff <- 0.01
# log2fc_cutoff <- 3
# ntop <- 10
# contrasts <- c("clusterPanc", "clusterLn", "clusterKidney", "clusterColon", "clusterBrain")
# 
# g <- filter(de_qq_qn, contrast %in% contrasts,
#             abs(log2fc) > log2fc_cutoff, 
#             padj < padj_cutoff) %>%
#   group_by(contrast, de) %>%
#   slice_max(abs(log2fc), n=ntop) %>%
#   ungroup() %>%
#   distinct(gene) %>% 
#   pull(gene)
# ngenes <- length(g)
# cluster_marker_genes <- g
# 
# # centered gene expression (subtract median)
# s <- fds_qq_qn$samples
# x <- log2_norm_cpm_qq_qn[g, s$aoi]
# u <- apply(x, 1, mean)
# x <- sweep(x, 1, u, FUN="-")
# xmin <- -6
# xmax <- 6
# 
# # heatmap
# heatmap_dir <- file.path(plot_dir, "heatmap")
# 
# f <- file.path(plot_dir,  paste0("heatmap_de_cluster_markers_", ngenes, "_genes"))
# h <- 6
# w <- 6
# p <- plot_pheatmap(x, 
#                    annot_col[s$aoi,], 
#                    annot_row[g,], 
#                    annot_colors=scale_color_continuous(),
#                    filename=paste0(f, ".png"), 
#                    width=w, height=h,
#                    xmin=xmin, xmax=xmax)
# save_pheatmap_pdf(p, filename=paste0(f, ".pdf"), width=w, height=h)

@@

for (gene in feature_plot_genes) {
  x <- bind_cols(s, as_tibble(pca_res_qq_qn$x), gene=unlist(y[gene,]))
  p <- ggplot(x, aes(x=PC1, y=PC2, color=gene, shape=study)) +
    geom_point(alpha = 0.8, size=3) +
    scale_color_viridis_c() +
    theme_minimal() +
    theme(aspect.ratio=1) +
    labs(x="PC1", y="PC2", title=gene)
  f <- file.path(plot_dir, paste0("pca_featureplot_", gene))
  ggsave(paste0(f, ".pdf"), plot=p, width=5, height=5)
  ggsave(paste0(f, ".png"), plot=p, width=5, height=5)
}

@@

# plot_de_scatter2d <- function(tbl, xcontrast, ycontrast, nlabel=10, 
#                               highlight_genes=NULL) {
#   if (is.null(highlight_genes)) {
#     highlight_genes <- c()
#   }
#   xde <- sym(paste0("de_", xcontrast))
#   yde <- sym(paste0("de_", ycontrast))
#   xlog2fc <- sym(paste0("log2fc_", xcontrast))
#   ylog2fc <- sym(paste0("log2fc_", ycontrast))
# 
#   x <- tbl %>% 
#     mutate(subtype = paste0("x", {{xde}}, "_", "y", {{yde}}),
#            log2fc_abs_x = case_when({{xde}} == "no" ~ 0,
#                                     TRUE ~ abs({{xlog2fc}})),
#            log2fc_abs_y = case_when({{yde}} == "no" ~ 0,
#                                     TRUE ~ abs({{ylog2fc}})),
#            log2fc_abs_max = pmax(log2fc_abs_x, log2fc_abs_y))
#   
#   color_scale = c(xno_yno = "#888888",
#                   xup_yup = "#FF0000",
#                   xdn_ydn = "#0000aa",
#                   xup_ydn = "#eeee00",
#                   xdn_yup = "#ff00ff",
#                   xup_yno = "#32cd32",
#                   xdn_yno = "#670092",
#                   xno_yup = "#ff6700",
#                   xno_ydn = "#16ccc4")
#   size_scale = c(xno_yno = 1, 
#                  xup_yup = 2,
#                  xdn_ydn = 2,
#                  xup_ydn = 2,
#                  xdn_yup = 2,
#                  xup_yno = 2,
#                  xdn_yno = 2,
#                  xno_yup = 2,
#                  xno_ydn = 2)
#   
#   annot_data <- filter(x, subtype != "xno_yno")
#   none_data <- filter(x, subtype == "xno_yno")
#   label_data <- annot_data %>% 
#     group_by(subtype) %>%
#     slice_max(log2fc_abs_max, n=nlabel) %>%
#     ungroup()
#   highlight_data <- filter(x, gene %in% highlight_genes)
# 
#   p <- ggplot(x, aes(x={{xlog2fc}}, 
#                      y={{ylog2fc}},
#                      color=subtype, 
#                      size=subtype)) +
#     geom_point(data=none_data, alpha=0.4) + 
#     geom_point(data=annot_data, alpha=0.8) +
#     geom_text_repel(data=label_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) + 
#     geom_point(data=highlight_data, pch=21, fill=NA, size=5, colour="#ffff00", stroke=1) + 
#     geom_point(data=highlight_data, pch=21, fill=NA, size=4, colour="#222222", stroke=1, alpha=0.5) +
#     geom_text_repel(data=highlight_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) +     
#     scale_color_manual(values = color_scale) +
#     scale_size_manual(values = size_scale, guide = "none") + 
#     theme_minimal() +
#     xlab(rlang::as_string(xlog2fc)) +
#     ylab(rlang::as_string(ylog2fc)) +
#     labs(color = "DE Genes")
#   return(p)
# }
# 
# 
# # merge de analysis by gene
# de_merged <- de_qq_qn %>% select(gene, contrast, log2fc, padj, de) %>%
#   pivot_wider(names_from = contrast, values_from = c(log2fc, padj, de))
# nlabel <- 10
# width <- 8
# height <- 8
# 
# #
# # IPMN Clusters
# #
# p <- plot_de_scatter2d(de_merged, 
#                        xcontrast="clusterColon", 
#                        ycontrast="clusterLn", 
#                        nlabel=nlabel) +
#   labs(title="IPMN Specimens", 
#        subtitle="clusterColon (x-axis) clusterLn (y-axis)")
# f <- file.path(plot_dir, "scatter2d_de_x_clusterColon_y_clusterLn")
# ggsave(paste0(f, ".pdf"), plot=p, width=8, height=8)
# ggsave(paste0(f, ".png"), plot=p, width=8, height=8)

@@

## FIGURE 6: GSEA
``` {r gsea}
# BiocManager::install("fgsea")
# 
# gsea_nperms <- 10000
# 
# read_msigdb_json_file <- function(filename) {
#   x <- jsonlite::fromJSON(filename)
#   y <- tibble(gs_name = names(x), gs = x) %>% 
#     unnest_wider(gs) %>%
#     unnest_longer(geneSymbols) %>%
#     select(gs_cat = collection,
#            gs_name = gs_name,
#            gene_symbol = geneSymbols)
# }
# 
# # read msigdb data: hallmark
# f <- system.file("extdata", "h.all.v2024.1.Hs.json", package = "stgeomx")
# msigdb_gene_sets <- read_msigdb_json_file(f)
# # determine gene universe
# msigdb_genes <- intersect(fds_qq_qn$meta$gene, msigdb_gene_sets$gene_symbol)
# # filter genes not in our dataset
# msigdb_gene_sets <- msigdb_gene_sets %>%
#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# # gene set categories
# gs_hallmark <- dplyr::filter(msigdb_gene_sets, gs_cat == "H")
# gs_hallmark <- split(x = gs_hallmark$gene_symbol, f = gs_hallmark$gs_name)
# 
# # read msigdb data: pancreas_endocrine
# f <- system.file("extdata", "GOBP_ENDOCRINE_PANCREAS_DEVELOPMENT.v2024.1.Hs.json", package = "stgeomx")
# gobp_end_pancreas_gene_sets <- read_msigdb_json_file(f)
# # determine gene universe
# gobp_end_pancreas_genes <- intersect(fds_qq_qn$meta$gene, gobp_end_pancreas_gene_sets$gene_symbol)
# # filter genes not in our dataset
# gobp_end_pancreas_gene_sets <- gobp_end_pancreas_gene_sets %>%
#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# # gene set categories
# gs_end_panc <- dplyr::filter(gobp_end_pancreas_gene_sets, gs_cat == "C5:GO:BP")
# gs_end_panc <- split(x = gs_end_panc$gene_symbol, f = gs_end_panc$gs_name)
# 
# # read msigdb data: pancreas secretion
# f <- system.file("extdata", "GOBP_INSULIN_SECRETION.v2024.1.Hs.json", package = "stgeomx")
# gobp_pancreas_secretion_gene_sets <- read_msigdb_json_file(f)
# # determine gene universe
# gobp_pancreas_secretion_genes <- intersect(fds_qq_qn$meta$gene, gobp_pancreas_secretion_gene_sets$gene_symbol)
# # filter genes not in our dataset
# gobp_pancreas_secretion_gene_sets <- gobp_pancreas_secretion_gene_sets %>%
#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# # gene set categories
# gs_panc_sec <- dplyr::filter(gobp_pancreas_secretion_gene_sets, gs_cat == "C5:GO:BP")
# gs_panc_sec <- split(x = gs_panc_sec$gene_symbol, f = gs_panc_sec$gs_name)
# 
# # read msigdb data: brain development gene sets
# f <- system.file("extdata", "brain_genesets.v2024.1.Hs.json", package = "stgeomx")
# gobp_brain_gene_sets <- read_msigdb_json_file(f)
# # determine gene universe
# gobp_brain_genes <- intersect(fds_qq_qn$meta$gene, gobp_brain_gene_sets$gene_symbol)
# # filter genes not in our dataset
# gobp_brain_gene_sets <- gobp_brain_gene_sets %>%
#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# # gene set categories
# gs_brain <- dplyr::filter(gobp_brain_gene_sets, gs_cat == "C5:GO:BP")
# gs_brain <- split(x = gs_brain$gene_symbol, f = gs_brain$gs_name)
# 
# # read msigdb data: ln development gene sets
# f <- system.file("extdata", "ln_genesets.v2024.1.Hs.json", package = "stgeomx")
# gobp_ln_gene_sets <- read_msigdb_json_file(f)
# # determine gene universe
# gobp_ln_genes <- intersect(fds_qq_qn$meta$gene, gobp_ln_gene_sets$gene_symbol)
# # filter genes not in our dataset
# gobp_ln_gene_sets <- gobp_ln_gene_sets %>%
#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# # gene set categories
# gs_ln <- dplyr::filter(gobp_ln_gene_sets, gs_cat == "C5:GO:BP")
# gs_ln <- split(x = gs_ln$gene_symbol, f = gs_ln$gs_name)
# 
# # read msigdb data: kidney development gene sets
# f <- system.file("extdata", "kidney_genesets.v2024.1.Hs.json", package = "stgeomx")
# gobp_kidney_gene_sets <- read_msigdb_json_file(f)
# # determine gene universe
# gobp_kidney_genes <- intersect(fds_qq_qn$meta$gene, gobp_kidney_gene_sets$gene_symbol)
# # filter genes not in our dataset
# gobp_kidney_gene_sets <- gobp_kidney_gene_sets %>%
#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# # gene set categories
# gs_kidney <- dplyr::filter(gobp_kidney_gene_sets, gs_cat == "C5:GO:BP")
# gs_kidney <- split(x = gs_kidney$gene_symbol, f = gs_kidney$gs_name)
# 
# # read msigdb data: colon development gene sets
# f <- system.file("extdata", "colon_genesets.v2024.1.Hs.json", package = "stgeomx")
# gobp_colon_gene_sets <- read_msigdb_json_file(f)
# # determine gene universe
# gobp_colon_genes <- intersect(fds_qq_qn$meta$gene, gobp_colon_gene_sets$gene_symbol)
# # filter genes not in our dataset
# gobp_colon_gene_sets <- gobp_colon_gene_sets %>%
#   dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
#   distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# # gene set categories
# gs_colon <- dplyr::filter(gobp_colon_gene_sets, gs_cat == "C5:GO:BP")
# gs_colon <- split(x = gs_colon$gene_symbol, f = gs_colon$gs_name)
# 
# # new gsea code
# get_fgsea_ranks_de <- function(xde, mycontrast) {
#   ranks <- xde %>%
#     dplyr::filter(contrast == mycontrast) %>%
#     select(gene, log2fc, padj) %>%
#     mutate(rank = log2fc)
#   ranks = sort(setNames(ranks$rank, ranks$gene), decreasing = TRUE)
#   return(ranks)
# }
#  
#  
# fgsea_run_batch <- function(xde, xgs,
#                             fgsea_eps=0,
#                             fgsea_nperms=1000) {
#   gsea <- NULL
#   for (xcontrast in unique(xde$contrast)) {
#     ranks <- get_fgsea_ranks_de(xde, xcontrast)
#     res <- fgsea(pathways=xgs,
#                  stats=ranks,
#                  eps=fgsea_eps,
#                  nPermSimple=fgsea_nperms)
#     res <- mutate(res, contrast = xcontrast)
#     gsea <- bind_rows(gsea, res)     
#   }
#   return(gsea)
# }
# 
# stgeomx_gsea_pipeline <- function(gsea, de, method) {
#   res <- fgsea_run_batch(de, gs_hallmark,
#                        fgsea_eps=0, fgsea_nperms=gsea_nperms)
#   res <- res %>% mutate(method = method)
#   gsea <- bind_rows(gsea, res)
#   
#   res <- fgsea_run_batch(de, gs_ex_panc,
#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
#   res <- res %>% mutate(method = method)
#   gsea <- bind_rows(gsea, res)
#   
#   
#   res <- fgsea_run_batch(de, gs_end_panc,
#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
#   res <- res %>% mutate(method = method)
#   gsea <- bind_rows(gsea, res)
#   
#   res <- fgsea_run_batch(de, gs_panc_sec,
#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
#   res <- res %>% mutate(method = method)
#   gsea <- bind_rows(gsea, res)
#   
#   res <- fgsea_run_batch(de, gs_brain,
#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
#   res <- res %>% mutate(method = method)
#   gsea <- bind_rows(gsea, res)
#   
#   res <- fgsea_run_batch(de, gs_ln,
#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
#   res <- res %>% mutate(method = method)
#   gsea <- bind_rows(gsea, res)
#   
#   res <- fgsea_run_batch(de, gs_kidney,
#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
#   res <- res %>% mutate(method = method)
#   gsea <- bind_rows(gsea, res)
#   
#   res <- fgsea_run_batch(de, gs_colon,
#                          fgsea_eps=0, fgsea_nperms=gsea_nperms)
#   res <- res %>% mutate(method = method)
#   gsea <- bind_rows(gsea, res)
#   
#   # filter out irrelevant genes in hallmark
#   gsea <- gsea %>%
